---
title: "Exploring NCAA Referee and Game Trends with `bbinsightsR`"
author: "Maya Parker, Bradford Rasmussen, Dallin Wilson"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Using the bbinsightsR Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

The `bbinsightsR` package provides tools for analyzing NCAA basketball play-by-play data with a focus on referee behavior, momentum changes, and free throw accuracy. This is useful for fans, analysts, and sports statisticians who want to investigate officiating fairness or team performance patterns.

A college basketball game is typically officiated by **three referees**, who rotate between lead, center, and trail positions. Referees have a major impact: if a player accumulates **five fouls**, they are disqualified from the game, so inconsistency in officiating can significantly affect outcomes.

**Free throws** are awarded for fouls and are taken without defensive pressure — often in high-pressure moments — and are especially influenced by the game environment (e.g., noisy crowds at home arenas).


> GitHub Repository: <https://github.com/mayaparker/bbinsightsr>

# Installation

Install the development version from GitHub using `pak`:

``` {r, eval=FALSE, echo=TRUE}
install.packages("pak")
pak::pak("mayaparker/bbinsightsr")
```

# Load the Package

``` {r}
library(bbinsightsR)
```

# Dataset Overview

The package includes `usu_data`, a dataset containing five seasons of play-by-play data for Utah State men’s basketball games. 

``` {r}
head(usu_data)
```
Each row represents an event, such as a shot, foul, or free throw, and includes metadata like the team, time, and player involved.

# Referee Analysis

## Generate Referee Statistics

You can use `refstats()` to analyze individual referee behavior based on fouls and calls involving a specific team:

``` {r}
stats <- refstats(usu_data, team_name = "Utah State")
head(stats)
```

This shows how balanced or biased a referee may be towards a team as well as overall fouls called. A foul ratio near 1 indicates neutrality, while a skewed value may warrant closer review. 

## Visualize Referee Tendencies

The `plot_refstats()` function can be used to visualize the most active referees and how they interact with your team of interest:

``` {r}
plot_refstats(stats, top_n = 5, team_label = "Utah State")
```

These bar charts show the five referees with the most fouls in Utah State games, most fouls against each team, and highest fouls ratios against Utah State. 

# Free Throw Analysis

The free throw analysis functions provide insights in free throws made during the first vs. second half. This can inform a team if the student section impacts free throws made when facing opposing teams.

## Opponent Comparison

The ft_opponent function produces a side-by-side free throw percentage comparison by half for each team.

``` {r}
ft_opponent(
  df = usu_data,
  home_team = "Utah State",
  away_team = "Boise St",
  arena = "Dee Glen Smith Spectrum"
)
```

This shows the free throw summary between Utah State and Boise St. The t-test shows that there is not a significant difference between the free throws made in each half for Boise st however it is significant for Utah State. 

## Arena-Based Analysis

The ft_stadium function analyzes cumulative free throw percentages for home and away teams in a specific arena over many games.

``` {r}
ft_stadium(
  df = usu_data,
  home_team = "Utah State",
  arena = "Dee Glen Smith Spectrum"
)
```

This example shows average free throws made for the 1st half and second half in the Utah State home stadium. It shows that there is not a significant difference between them. 

# Game Trends

The `plot_game_trends()` plots a desired statistic across a desired segment of the game, by breaking the game up between timeouts, halves, or splitting the game into a specific number of segments. This shows the trends of the desired statistic throughout the game.

``` {r fig.width = 9, fig.height = 5}
plot_game_trends(
  games      = usu_data,
  home_team  = "Utah State",
  away_team  = "Wyoming",
  game_break = "timeout",
  stat       = "points",
  game_date  = "2023-01-10"
)
```
This example plots the number of points between each timeout for the Utah State vs. Wyoming game on January 10th 2023 allowing the user to look for trends and scoring throughout the game for each team. 

# Conclusion

The `bbinsightsR` package supports advanced play-by-play data analysis for NCAA basketball, offering new perspectives on referee patterns, free throw trends, and momentum swings. These tools can assist analysts, researchers, and fans in extracting meaningful insights from game data.

# About

This package was created by Maya Parker, Bradford Rasmussen, and Dallin Wilson for STAT 5555/6555 at Utah State University. It is licensed under GPL-3.

